---
title: "Descriptive Stats"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
```

# Read in and shape data  

Really should source another script or write a function to do this, but I'm in the early days.

```{r}
#### READ IN DATA ####
# working off NEveg outputs here; our template will be different
# these have cover, density, and height in the same sheet
# and an extra row at the top for which marsh zone each plant "belongs to"
# anything that does NOT have 'Canopy Ht' or 'Density' in the column name 
# once identifying info stops is percent cover
grb_in <- here::here("data", "GRB Data Packet 6-8-21.xlsx")
wqb_in <- here::here("data", "WQB Data Packet 2011-2020 kac_fixes.xlsx")
```

## Change reserve HERE

```{r}
# work with GRB first
# dat <- read_xlsx(grb_in, sheet = "Veg",
#                  skip = 1)
```

```{r}
# WQB - but have to get rid of extra columns -
# 264 to end
dat <- read_xlsx(wqb_in, sheet = "Veg",
                 skip = 1)
# dat <- dat[,1:263] # AVOID THIS. CONVINCE CHRIS ABOUT SEPARATING DIFFERENT TYPES OF DATA FOR PREDICTABILITY.
```


```{r}
# bust up into data types, like we'll get in the template
id_cols <- dat %>% 
    select("Reserve Code":"Marsh Zone")
density <- dat %>% 
    select(contains("Density"))
height <- dat %>% 
    select(contains("Canopy Ht"))
cover <- dat %>% 
    select(-("Reserve Code":"Marsh Zone"),
           -contains("Density"),
           -contains("Canopy Ht"))

# make sure the numbers add up
ncol(dat) == ncol(id_cols) + ncol(density) + ncol(height) + ncol(cover)

# for now, just work with cover
# WQB has a 0,5 instead of 0.5 in the Toxicodendron radicans column
# FIGURE OUT A WAY TO FIND THIS TYPE OF THING BEFORE CRUNCHING NUMBERS
cover <- cover %>% 
    select(-contains("Unknown"),
           -contains("Notes")) %>% 
    mutate_all(as.numeric) %>% 
    cbind(id_cols, .)

# pivot to long
cover_long <- pivot_longer(cover,
                           cols = (ncol(id_cols) + 1):ncol(cover),
                           names_to = "Species",
                           values_to = "pct_cover") %>% 
    janitor::clean_names()

# cringe to do this, but turn NAs into 0s, because if all data was entered properly,
# anything that wasn't recorded should be 0.
# will be important to check individual data points in some graphs
# to make sure there aren't any mistaken 0s.

cover_long <- cover_long %>% 
    mutate(pct_cover = case_when(is.na(pct_cover) ~ 0,
                                 TRUE ~ pct_cover))

```

# Descriptive info/metadata-ish stuff    

What do I want to know:  

+  how many sites are there?  
+  how many transects per site?  
+  how many plots per transect?  
+  date range?  
+  missing data?  

Would be good to turn this into a shiny app or something; it's going to be a LOT of output otherwise  

```{r}
cover_long %>% 
    summarize(n_sites = length(unique(site)),
              date_min = min(date),
              date_max = max(date),
              n_plots = length(unique(plot_id)),
              n_species = length(unique(species))) %>% 
    knitr::kable()
```

```{r}
cover_long %>% 
    group_by(site) %>% 
    summarize(date_min = min(date),
              date_max = max(date),
              n_transects = length(unique(transect_number)),
              n_plots = length(unique(plot_id)),
              n_species = length(unique(species))) %>% 
    knitr::kable()
```

All species:  

```{r}
unique(cover_long$species)
```

How to count species where most aren't NA?

```{r}
# get rid of totals
to_exclude <- c("TOTAL", "TOTAL W/O Water")
```

```{r}
non_zeros <- cover_long %>% 
    group_by(site, species) %>% 
    filter(!(species %in% to_exclude)) %>% 
    summarize(non_zeros = sum(pct_cover > 0),
              zeros = sum(pct_cover == 0)) %>%
    mutate(pct_non_zeros = round(non_zeros*100/(non_zeros + zeros), 1)) %>% 
    arrange(desc(pct_non_zeros))

knitr::kable(non_zeros[1:100, ])

```


## NEED: Build in flexibility for species selection  



# Exploratory analysis of vegetation data  

Here I want to know things like:  

+  what is the average percent cover by species and transect  
+  how does that look over time: showing all plots, averaging to transect, maybe averaging to station?  
+  how many species appear  
+  how many species appear a LOT (unsure how to define this for R; it's a "know it when I see it" kind of thing) - what's dominant, what's common, etc. because we can potentially narrow everything down to just a few species.  
+  NE project made pie charts for each site, each year. Could do this. Potentially also stacked bar chart or something. Unsure what would be cleanest.  
  +  probably have to combine some (several) species into an "other" category. `dplyr` has some way to do this, right? anything making up less than X% gets lumped together?   
    +  `forcats::fct_lump()` is it. source: https://stackoverflow.com/a/55093206   
    +  help file for that led me to `forcats::fct_other`, which lets you specify a vector of values to keep/drop. Given what I've done below, this seems most useful.  
  +  can facet a stacked bar chart by site. x = year, y = species (proportion at site), facet = site.  
  +  would a scatter/line plot work just as well? Maybe? could be easier to follow an individual species.....  




## Top_x_pct  

### Specify here:  

The below identifies species that were present more than `x`% of the time, at one or more sites. Alternately, could just choose some number of species?  

```{r}
pct <- 25
```

Let R do the rest:  

```{r}
x_pct <- non_zeros %>% 
    filter(pct_non_zeros >= pct) %>% 
    ungroup() %>% 
    summarize(species = unique(species)) %>% 
    arrange(species)

nrow(x_pct)
knitr::kable(x_pct, caption = paste0("Species that show up at least ", pct, "% of the time at one or more sites"))
```

```{r}
# get rid of totals
to_exclude <- c("TOTAL", "TOTAL W/O Water")

# trim and create "Other" category
trimmed <- cover_long %>% 
    filter(!(species %in% to_exclude)) %>% 
    mutate(species = fct_other(species, keep = c(x_pct$species)))

# add up all the "Other"s
trimmed2 <- trimmed %>% 
    group_by(site, year, date, transect_number, plot_id, species) %>% 
    summarize(pct_cover = sum(pct_cover))
```

## Plotting function  

```{r}
plot_one_site <- function(data, site, ncol, by_transect = FALSE){
    dat <- data[which(data$site == site), ]
    if(by_transect == TRUE){
      dat <- dat %>% 
        group_by(site, year, date, transect_number, species) %>% 
        summarize(pct_cover = mean(pct_cover))
    }
    ggplot(dat) +
        geom_point(aes(x = year, y = pct_cover,
                       color = as.factor(transect_number)),
                   alpha = 0.5, size = 2) +
        facet_wrap(~species, ncol = ncol) +
        labs(title = site,
             color = "Transect") +
        theme(legend.position = "right")
}
# data = data frame
# site = IN QUOTES, the site you want to see
# ncol = number of columns
```

```{r}
# plot_one_site(trimmed2, "Bunker Creek", 3)
# plot_one_site(trimmed2, "Bunker Creek", 3, by_transect = TRUE)
# plot_one_site(trimmed2, "Great Bay Farms", 3)

all_sites <- unique(trimmed2$site)
for(i in seq_along(all_sites)){
    print(plot_one_site(trimmed2, all_sites[i], 3, by_transect = TRUE))
}
```


